---
title: "Louisiana Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Louisiana Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(laschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6")
```

```{r fetch-data, cache = TRUE}
enr <- fetch_enr_multi(2019:2024, use_cache = TRUE)
enr_current <- fetch_enr(2024, use_cache = TRUE)

# Calculate state totals for percentage calculations
state_totals <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, total = n_students)
```

## 1. Louisiana added 33,000 students since 2019

Statewide enrollment grew 5.1% from 2019 to 2024, rising from 644,000 to 677,000 students.

```{r state-growth}
state_trend <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(end_year, n_students)
stopifnot(nrow(state_trend) > 0)
state_trend

ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2021.1, y = max(state_trend$n_students), label = "COVID",
           hjust = 0, color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Louisiana Statewide Enrollment",
       subtitle = "5.1% growth from 2019 to 2024 (676,751 students)",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 2. White students fell below Black students for the first time

White enrollment dropped from 47.1% in 2019 to 40.7% in 2024, while Black students held steady at 41.7%, making Louisiana's public schools majority-minority.

```{r demographics}
demo <- enr %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  left_join(state_totals, by = "end_year") %>%
  mutate(pct = n_students / total * 100)
stopifnot(nrow(demo) > 0)
demo %>% filter(end_year == 2024) %>% select(subgroup, n_students, pct)

ggplot(demo, aes(x = end_year, y = pct, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = colors,
                     labels = c("Asian", "Black", "Hispanic", "White")) +
  labs(title = "Louisiana Demographics",
       subtitle = "White share fell from 47% to 41%, now below Black share",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_readme()
```

## 3. Hispanic enrollment surged 45% in five years

Hispanic students grew from 53,778 (8.4%) in 2019 to 77,836 (11.5%) in 2024, the fastest-growing demographic group in Louisiana schools.

```{r hispanic-surge}
hisp <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "hispanic") %>%
  left_join(state_totals, by = "end_year") %>%
  mutate(pct = n_students / total * 100)
stopifnot(nrow(hisp) > 0)
hisp %>% select(end_year, n_students, pct)

ggplot(hisp, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["hispanic"]) +
  geom_point(size = 3, color = colors["hispanic"]) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Hispanic Enrollment Surging in Louisiana",
       subtitle = "45% growth from 53,778 to 77,836 students (2019-2024)",
       x = "School Year", y = "Hispanic Students") +
  theme_readme()
```

## 4. Jefferson Parish is Louisiana's largest district

Jefferson Parish enrolls 47,702 students, leading all districts and outpacing East Baton Rouge (39,932) and St. Tammany (36,384).

```{r top-districts}
top5 <- enr_current %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_name != "State of Louisiana") %>%
  arrange(desc(n_students)) %>%
  head(5)
stopifnot(nrow(top5) > 0)
top5 %>% select(district_name, n_students)

top5_names <- top5$district_name

top5_trend <- enr %>%
  filter(is_district, district_name %in% top5_names,
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(top5_trend) > 0)

ggplot(top5_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Louisiana's Five Largest Districts",
       subtitle = "Jefferson Parish leads with 47,702 students in 2024",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 5. Caddo Parish lost 14% of its students in five years

Caddo Parish (Shreveport) dropped from 37,868 students in 2019 to 32,614 in 2024, a loss of over 5,200 students.

```{r caddo-decline}
caddo <- enr %>%
  filter(is_district, district_name == "Caddo Parish",
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(caddo) > 0)
caddo %>% select(end_year, district_name, n_students)

ggplot(caddo, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Caddo Parish (Shreveport) Enrollment Decline",
       subtitle = "Lost 5,254 students (-13.9%) from 2019 to 2024",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 6. COVID wiped out 17% of Pre-K enrollment overnight

Pre-K dropped from 26,078 to 21,751 students between 2019 and 2020, then slowly recovered to 26,152 by 2024.

```{r prek-covid}
prek <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "PK")
stopifnot(nrow(prek) > 0)
prek %>% select(end_year, n_students)

ggplot(prek, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020.6, y = max(prek$n_students), label = "COVID",
           hjust = 0, color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Pre-K Enrollment Cratered During COVID",
       subtitle = "-16.6% drop in 2020, full recovery by 2024",
       x = "School Year", y = "Pre-K Students") +
  theme_readme()
```

## 7. Kindergarten also took a COVID hit

Kindergarten enrollment fell 6.9% from 48,556 to 45,205 between 2019 and 2020, and still has not returned to pre-pandemic levels.

```{r covid-kindergarten}
k_trend <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("PK", "K", "01", "09")) %>%
  mutate(grade_label = case_when(
    grade_level == "PK" ~ "Pre-K",
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "09" ~ "Grade 9"
  ))
stopifnot(nrow(k_trend) > 0)
k_trend %>%
  filter(grade_level == "K") %>%
  select(end_year, grade_label, n_students)

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020.6, y = max(k_trend$n_students), label = "COVID",
           hjust = 0, color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID Impact by Grade Level",
       subtitle = "Youngest grades hit hardest, K still below pre-pandemic levels",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 8. English learners grew from 3.9% to 5.3% of enrollment

LEP students increased from 24,908 to 35,868 between 2019 and 2024, a 44% jump tracking closely with Hispanic growth.

```{r el-growth}
el <- enr %>%
  filter(is_state, subgroup == "lep", grade_level == "TOTAL") %>%
  left_join(state_totals, by = "end_year") %>%
  mutate(pct = n_students / total * 100)
stopifnot(nrow(el) > 0)
el %>% select(end_year, n_students, pct)

ggplot(el, aes(x = end_year, y = pct)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  labs(title = "English Learners on the Rise",
       subtitle = "From 3.9% to 5.3% of enrollment (2019-2024)",
       x = "School Year", y = "Percent of Students") +
  theme_readme()
```

## 9. Seven in ten Louisiana students are economically disadvantaged

The statewide economic disadvantage rate has hovered around 70% since 2019, one of the highest rates in the nation.

```{r econ-state}
econ_state <- enr %>%
  filter(is_state, subgroup == "econ_disadv", grade_level == "TOTAL") %>%
  left_join(state_totals, by = "end_year") %>%
  mutate(pct = n_students / total * 100)
stopifnot(nrow(econ_state) > 0)
econ_state %>% select(end_year, n_students, pct)

ggplot(econ_state, aes(x = end_year, y = pct)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(limits = c(60, 80)) +
  labs(title = "Economic Disadvantage Rate in Louisiana",
       subtitle = "Consistently around 70% statewide",
       x = "School Year", y = "Percent Economically Disadvantaged") +
  theme_readme()
```

## 10. St. Helena Parish: 100% economically disadvantaged

St. Helena leads Louisiana with every single student classified as economically disadvantaged, followed by East Carroll (96.9%) and Tensas (95.8%).

```{r econ-disadvantage}
district_totals <- enr_current %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(district_name, total = n_students)

econ <- enr_current %>%
  filter(is_district, subgroup == "econ_disadv", grade_level == "TOTAL") %>%
  left_join(district_totals, by = "district_name") %>%
  mutate(pct = n_students / total * 100) %>%
  arrange(desc(pct)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, pct))
stopifnot(nrow(econ) > 0)
econ %>% select(district_name, n_students, total, pct)

ggplot(econ, aes(x = district_label, y = pct)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  labs(title = "Highest Poverty Districts in Louisiana",
       subtitle = "St. Helena Parish reports 100% economically disadvantaged",
       x = "", y = "Percent Economically Disadvantaged") +
  theme_readme()
```

## 11. Delta parishes are emptying out

East Carroll (-20.8%), Tensas (-22.6%), and Madison (-4.4%) have shed students steadily since 2019.

```{r rural-decline}
delta_names <- c("Tensas Parish", "East Carroll Parish", "Madison Parish")
delta_trend <- enr %>%
  filter(is_district, district_name %in% delta_names,
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(delta_trend) > 0)
delta_trend %>% select(end_year, district_name, n_students)

delta_combined <- delta_trend %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")
delta_combined

ggplot(delta_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Delta Parishes Losing Students",
       subtitle = "Tensas, East Carroll, and Madison combined lost 13.4% since 2019",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 12. Calcasieu Parish never recovered from Hurricane Laura

Calcasieu Parish (Lake Charles) lost 11.3% of enrollment in a single year (2019-2020) after Hurricane Laura devastated southwest Louisiana.

```{r calcasieu-hurricane}
calcasieu <- enr %>%
  filter(is_district, district_name == "Calcasieu Parish",
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(calcasieu) > 0)
calcasieu %>% select(end_year, district_name, n_students)

ggplot(calcasieu, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020.1, y = 31500, label = "Hurricane Laura",
           hjust = 0, color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Calcasieu Parish (Lake Charles)",
       subtitle = "Lost 3,600 students after Hurricane Laura, partial recovery",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 13. Jefferson Parish: the slow suburban slide

Louisiana's largest district has been quietly losing students, dropping from 50,566 in 2019 to 47,702 in 2024.

```{r jefferson-decline}
jefferson <- enr %>%
  filter(is_district, district_name == "Jefferson Parish",
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(jefferson) > 0)
jefferson %>% select(end_year, district_name, n_students)

ggplot(jefferson, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Jefferson Parish - Largest but Shrinking",
       subtitle = "Lost 2,864 students (-5.7%) from 2019 to 2024",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 14. The suburban corridor holds steady

Ascension, Livingston, and St. Tammany parishes along the I-10/I-12 corridor are among the few growing districts.

```{r i10-growth}
i10 <- c("Livingston Parish", "Ascension Parish", "St. Tammany Parish")
i10_trend <- enr %>%
  filter(is_district, district_name %in% i10,
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(i10_trend) > 0)
i10_trend %>%
  select(end_year, district_name, n_students) %>%
  tidyr::pivot_wider(names_from = district_name, values_from = n_students)

ggplot(i10_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "I-10/I-12 Corridor Parishes",
       subtitle = "Ascension and Livingston growing; St. Tammany stable",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 15. Gender balance is remarkably stable

Louisiana's 51.2% male / 48.8% female split has barely budged across all six years of data.

```{r gender-balance}
gender <- enr %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) %>%
  left_join(state_totals, by = "end_year") %>%
  mutate(pct = n_students / total * 100)
stopifnot(nrow(gender) > 0)
gender %>%
  filter(end_year == 2024) %>%
  select(subgroup, n_students, pct)

ggplot(gender, aes(x = end_year, y = pct, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("male" = "#3498DB", "female" = "#E74C3C"),
                     labels = c("Female", "Male")) +
  scale_y_continuous(limits = c(45, 55)) +
  labs(title = "Gender Balance in Louisiana Schools",
       subtitle = "Consistently 51.2% male / 48.8% female since 2019",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_readme()
```

## Session Info

```{r session-info}
sessionInfo()
```
