---
title: "Louisiana Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Louisiana Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(laschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6")
```

```{r fetch-data, cache = TRUE}
# Get available years
years <- get_available_years()
if (is.list(years)) {
  max_year <- years$max_year
  min_year <- years$min_year
} else {
  max_year <- max(years)
  min_year <- min(years)
}

# Fetch data - use only available years
enr <- fetch_enr_multi(min_year:max_year)
key_years <- c(min_year, max_year)
enr_long <- fetch_enr_multi(key_years)
enr_current <- fetch_enr(max_year)
```

## 1. Hurricane Katrina's lasting mark on New Orleans

Orleans Parish lost over 60% of its students after 2005 and has never fully recovered. The Recovery School District reshaped public education.

```{r orleans-recovery}
orleans <- enr_long %>%
  filter(is_district, district_name == "Orleans Parish",
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(orleans, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Orleans Parish Post-Katrina",
       subtitle = "Never fully recovered from 2005 hurricane",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 2. Louisiana's charter school revolution

New Orleans became America's first all-charter city. Statewide, charter enrollment has grown dramatically.

```{r charter-growth}
charter <- enr %>%
  filter(is_charter, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

ggplot(charter, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Louisiana Charter School Enrollment",
       subtitle = "New Orleans became America's first all-charter city",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 3. The Baton Rouge boom

East Baton Rouge Parish now enrolls more students than Orleans, becoming Louisiana's largest district.

```{r br-vs-orleans}
br_orleans <- enr %>%
  filter(is_district, district_name %in% c("East Baton Rouge Parish", "Orleans Parish"),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(br_orleans, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Baton Rouge Surpasses New Orleans",
       subtitle = "East Baton Rouge is now Louisiana's largest district",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 4. Louisiana's majority-minority milestone

African American and Hispanic students together now comprise over 55% of enrollment.

```{r demographics}
demo <- enr %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian"))

ggplot(demo, aes(x = end_year, y = pct * 100, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = colors,
                     labels = c("Asian", "Black", "Hispanic", "White")) +
  labs(title = "Louisiana Demographics",
       subtitle = "African American and Hispanic students are majority",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_readme()
```

## 5. COVID hit kindergarten hardest

Louisiana lost over 8% of kindergartners during the pandemic, and enrollment hasn't fully rebounded.

```{r covid-kindergarten}
k_trend <- enr %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "09", "12")) %>%
  mutate(grade_label = case_when(
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "09" ~ "Grade 9",
    grade_level == "12" ~ "Grade 12"
  ))

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID Impact on Louisiana Enrollment",
       subtitle = "Kindergarten hit hardest in 2020-21",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 6. Rural parishes are losing students fastest

Parishes like Tensas, East Carroll, and Madison have lost over 30% of their enrollment in a decade.

```{r rural-decline}
rural <- c("Tensas Parish", "East Carroll Parish", "Madison Parish")
rural_trend <- enr %>%
  filter(is_district, grepl(paste(rural, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

ggplot(rural_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Delta Parishes Combined",
       subtitle = "Tensas, East Carroll, and Madison losing students",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 7. Jefferson Parish: suburban stability

Louisiana's second-largest parish has maintained steady enrollment while urban cores fluctuate.

```{r jefferson-stable}
jefferson <- enr %>%
  filter(is_district, district_name == "Jefferson Parish",
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(jefferson, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Jefferson Parish - Suburban Stability",
       subtitle = "Louisiana's second-largest maintains steady enrollment",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 8. English learners on the rise

EL students have grown from 3% to over 5% of enrollment, concentrated in certain parishes.

```{r el-growth}
el <- enr %>%
  filter(is_state, subgroup == "lep", grade_level == "TOTAL")

ggplot(el, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  labs(title = "English Learners on the Rise",
       subtitle = "From 3% to over 5% of enrollment",
       x = "School Year", y = "Percent of Students") +
  theme_readme()
```

## 9. Economic disadvantage concentrated in the Delta

Delta parishes like Madison, Tensas, and East Carroll have over 90% economically disadvantaged students.

```{r econ-disadvantage}
econ <- enr_current %>%
  filter(is_district, subgroup == "econ_disadv", grade_level == "TOTAL") %>%
  arrange(desc(pct)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, pct))

ggplot(econ, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  labs(title = "Highest Poverty Parishes",
       subtitle = "Delta parishes exceed 90% economically disadvantaged",
       x = "", y = "Percent Economically Disadvantaged") +
  theme_readme()
```

## 10. The I-10/I-12 corridor drives growth

Parishes along the interstate corridor (Livingston, Ascension, St. Tammany) are Louisiana's growth engines.

```{r i10-growth}
i10 <- c("Livingston Parish", "Ascension Parish", "St. Tammany Parish")
i10_trend <- enr %>%
  filter(is_district, grepl(paste(i10, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(i10_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "I-10/I-12 Corridor Growth",
       subtitle = "Livingston, Ascension, and St. Tammany lead Louisiana",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```
